EGGs are adhesive!\\ An algebraic account of equality saturation

### 2025-09-09 jlamp/jlamp.tex
minima

\begin{prop}\label{prop:slice}
$\EqHyps$ is equivalent to $\EqHyp \downarrow R(\mathcal{G})$.
\end{prop}

@@

\begin{prop}\label{prop:varie}
	Let $\mathcal{G}$ be an hypergraph and $\Gamma_{R(\mathcal{G})}$ the costant functor in $R(\mathcal{G})$. Then the following facts hold
	\begin{enumerate}
		\item $\eggl$ is equivalent to $\comma{E}/\Gamma_{R(\mathcal{G})}$;
		\item there exists a functor $W_\mathcal{G}\colon \eggl\to \egg$ forgetting the labeling which creates all colimits, pullbacks and equalizers. 
	\end{enumerate}
\end{prop}

@@

\begin{cor}
	$\lelm$ is $\epbl_{\mathcal{G}}$-adhesive, while $\leps$ is $\epbs_{\mathcal{G}}$-adhesive.
\end{cor}

@@

	\noindent 
	\begin{minipage}[l]{.7\linewidth}
	Thus for each $d$ in $\D$, there exists an arrow $a_d:U\to U_i$ making the diagram on the right commutative.
	Then we have:
	\begin{align*}
		&l_d^\star\circ m^\star \circ q^\star \circ t \circ u_1    = q_d^\star\circ (\pi_V^d)^\star \circ t \circ u_1 \\
		= &q_d^\star \circ t_d \circ \pi_E^d \circ u_1
		= q_d^\star \circ t_d \circ u_1^d \circ a_d 
		= q_d^\star \circ t_d \circ u_2^d \circ a_d\\ 
		=& q_d^\star \circ t_d \circ \pi_E^d \circ u_2 = q_d^\star \circ (\pi_V^d)^\star \circ t \circ u_2 = l_d^\star \circ m^\star \circ q^\star \circ t \circ u_2
	\end{align*}
	\end{minipage}
\hfill
\begin{minipage}[r]{.3\linewidth}
\xymatrix@R=20pt{
	U \ar[r]^{u_1} \ar[d]_{u_2} \ar@{.>}[dr]_{a_d} & A \ar[dr]^{\pi_E^d} & \\
	A \ar[dr]_{\pi_E^d} & U_d \ar[r]^{u_1^d} \ar[d]_{u_2^d}& A_d \ar[d]^{q_d^\star \circ s_d} \\
	& A_i  \ar[r]_{q_d^\star \circ s_d}& Q_d^\star
}
\end{minipage}

	By universal property of limits, we have that \- $m^\star\circ q^\star \circ t \circ u_1 = m^\star \circ q^\star \circ t \circ u_2$. Since $m$ is mono we deduce that $q^\star \circ t \circ u_1 = q^\star \circ t \circ u_2$, hence the claim.
\end{proof}

\noindent
\begin{minipage}[l]{.7\linewidth}
\begin{rem}\label{rem:kleene}
	Let $\delta_1\colon 1\to \mathbb{N}$ be the arrow which picks $1\in \mathbb{N}$. Let $f\colon X\to Y$ be an arrow and consider theinclusion $v_1\colon X\to X^\star $  defined by the first point of \Cref{prop:fact}. Then we can build the cube aside which, by extensivity, has pullbacks as back, front and right faces. Thus, by \Cref{lem:pb1} its left face is a pullback too.
	\end{rem}
\end{minipage}
\hfill
\begin{minipage}[r]{.45\linewidth}
\xymatrix@C=15pt@R=5pt{&X\ar[dd]|\hole_(.65){!_{X}}\ar@{->}[rr]_{!_X} \ar@{>}[dl]_{f} && 1 \ar[dd]^{v_1} \ar[dl]^{\id{1}} \\ Y  \ar[dd]_{!_Y}\ar@{>}[rr]^(.7){!_Y} & & 1 \ar[dd]_(.3){\delta_1}\\&X^\star\ar@{>->}[rr]|(.53)\hole^(.70){\lgh_{X}} \ar@{>->}[dl]_{f^\star} && \mathbb{N} \ar[dl]^{\id{N}} \\Y^\star \ar@{>->}[rr]^(.7){\lgh_{Y}} & & \mathbb{N}}
\end{minipage}

\noindent
\begin{minipage}[l]{.83\linewidth}
\epo*
\end{minipage}
\hfill 
\begin{minipage}[r]{.15\linewidth}
	\xymatrix@R=10pt{\mathcal{G}_1 \ar[r]_{h}\ar@{>->}[d]_{m}&\mathcal{G}_2\ar[d]^{n}\\\mathcal{G}_3\ar[r]^{z}&\mathcal{P}}
\end{minipage}
\begin{proof}\label{proof:epo}
	Let $\mathcal{P} $ be $(A, B, C, s, t, q)$ and consider the kernel pair $(K_i, \pi_i^1, \pi_i^2)$ the kernel pair of $q_i^\star \circ s_i$, for $i\in \{1, 2, 3\}$. Let also $(U, u_1, u_2)$ be the kernel pair of $q^\star \circ s$.

	\vspace{-.3cm}
	\noindent
	\begin{minipage}[l]{.7\linewidth}
			\setlength{\parindent}{1.5em}
		Consider now the cube on the right. By hypothesis and \Cref{rem:kleene} its left face is a pullback, moreover its bottom face, while not a pushout it is still a pullback by the first point of \Cref{prop:regmono} and the third one of \Cref{prop:fact}.
	\end{minipage}\hfill 
	\begin{minipage}[r]{.4\linewidth}
	\xymatrix@C=10pt@R=2pt{&A_1\ar[dd]|\hole_(.65){q_1^\star \circ s_1}\ar[rr]^{h_E} \ar@{>->}[dl]_{m_E} && A_2 \ar[dd]^{q_2^\star \circ s_2} \ar@{>.>}[dl]_(.6){n_E} \\ A_3 \ar[dd]_{q_3^\star \circ s_3}\ar[rr]^(.7){z_E} & & A \ar[dd]_(.3){q^\star \circ s}\\&{Q_1^\star}\ar[rr]|\hole^(.65){h_Q^\star}\ar@{>->}[dl]_{m_Q^\star} && {Q_2^\star} \ar@{>->}[dl]^{n_Q^\star} \\{Q_3^\star} \ar[rr]^(.65){z_Q^\star} & & Q^\star}	\end{minipage}

	\noindent
	\begin{minipage}[l]{.2\linewidth}
		\xymatrix@R=15pt{K_1\ar[r]_{k_{h_E}}\ar@{>->}[d]_{k_{m_E}}&K_2\ar@{>->}[d]^{k_{n_E}}\\K_3\ar[r]^{k_{z_E}}&U}
	\end{minipage}\hfill
	\begin{minipage}[r]{.8\linewidth}
			\setlength{\parindent}{1.5em}
			By hypothesis and the first point of \Cref{prop:eqhyp_complete} the top face is a pushout. Thus we can apply \Cref{prop:kerset} to deduce that the square on the left is a pushout. 
			By computing we obtain:
	\end{minipage}
\vspace{-.3cm}
	\begin{align*}
		q^\star \circ t \circ u_1 \circ f_n &= q^\star \circ t \circ n_E \circ \pi_2^1 = n_C^\star \circ q_2^\star \circ s_2 \circ \pi_2^1 = n_C^\star \circ q_2^\star \circ s_2 \circ \pi_2^2 =q^\star \circ t \circ u_2 \circ f_n \\
		q^\star \circ t \circ u_1 \circ f_k &= q^\star \circ t \circ k_E \circ \pi_3^1 = k_C^\star \circ q_3^\star \circ s_3 \circ \pi_3^1 = k_C^\star \circ q_3^\star \circ s_3 \circ \pi_3^2 =q^\star \circ t \circ u_2 \circ f_k
	\end{align*}
	We can therefore deduce that $q^\star \circ t \circ u_1 = q^\star \circ t \circ u_2$, and the claim follows.
\end{proof}

>>>>

\begin{prop}\label{prop:slice}
$\EqHyps$ is equivalent to $\EqHyp / R(\mathcal{G})$.
\end{prop}

@@

\begin{prop}\label{prop:varie}
	Let $\mathcal{G}$ be an hypergraph and $\Gamma_{R(\mathcal{G})}$ the costant functor in $R(\mathcal{G})$. Then the following facts hold
	\begin{enumerate}
		\item $\eggl$ is equivalent to $\comma{E}\Gamma_{R(\mathcal{G})}$;
		\item the functor $W_\mathcal{G}\colon \eggl\to \egg$ forgetting the labeling creates all colimits, pullbacks and equalizers. 
	\end{enumerate}
\end{prop}

@@

\begin{cor}
	$\lelm$ is $\epbl_{\mathcal{G}}$-adhesive and $\leps$ is $\epbs_{\mathcal{G}}$-adhesive.
\end{cor}

@@

	\noindent 
	\begin{minipage}[l]{.7\linewidth}
	Thus for each $d$ in $\D$ there exists an arrow $a_d:U\to U_i$ making the diagram on the right commutative.
	Then we have
	\begin{align*}
		&l_d^\star\circ m^\star \circ q^\star \circ t \circ u_1    = q_d^\star\circ (\pi_V^d)^\star \circ t \circ u_1 \\
		= &q_d^\star \circ t_d \circ \pi_E^d \circ u_1
		= q_d^\star \circ t_d \circ u_1^d \circ a_d 
		= q_d^\star \circ t_d \circ u_2^d \circ a_d\\ 
		=& q_d^\star \circ t_d \circ \pi_E^d \circ u_2 = q_d^\star \circ (\pi_V^d)^\star \circ t \circ u_2 = l_d^\star \circ m^\star \circ q^\star \circ t \circ u_2
	\end{align*}
	\end{minipage}
\hfill
\begin{minipage}[r]{.3\linewidth}
\xymatrix@R=20pt{
	U \ar[r]^{u_1} \ar[d]_{u_2} \ar@{.>}[dr]_{a_d} & A \ar[dr]^{\pi_E^d} & \\
	A \ar[dr]_{\pi_E^d} & U_d \ar[r]^{u_1^d} \ar[d]_{u_2^d}& A_d \ar[d]^{q_d^\star \circ s_d} \\
	& A_i  \ar[r]_{q_d^\star \circ s_d}& Q_d^\star
}
\end{minipage}

	By the universal property of limits, we have that \- $m^\star\circ q^\star \circ t \circ u_1 = m^\star \circ q^\star \circ t \circ u_2$. Since $m$ is mono we deduce that $q^\star \circ t \circ u_1 = q^\star \circ t \circ u_2$, hence the claim.
\end{proof}

\noindent
\begin{minipage}[l]{.7\linewidth}
\begin{rem}\label{rem:kleene}
	Let $\delta_1\colon 1\to \mathbb{N}$ be the arrow which picks $1\in \mathbb{N}$. Let $f\colon X\to Y$ be an arrow and consider the inclusion $v_1\colon X\to X^\star $  defined by the first point of \Cref{prop:fact}. Then we can build the cube aside which, by extensivity, has pullbacks as back, front and right faces. Thus, by \Cref{lem:pb1} its left face is a pullback too.
	\end{rem}
\end{minipage}
\hfill
\begin{minipage}[r]{.45\linewidth}
\xymatrix@C=15pt@R=5pt{&X\ar[dd]|\hole_(.65){!_{X}}\ar@{->}[rr]_{!_X} \ar@{>}[dl]_{f} && 1 \ar[dd]^{v_1} \ar[dl]^{\id{1}} \\ Y  \ar[dd]_{!_Y}\ar@{>}[rr]^(.7){!_Y} & & 1 \ar[dd]_(.3){\delta_1}\\&X^\star\ar@{>->}[rr]|(.53)\hole^(.70){\lgh_{X}} \ar@{>->}[dl]_{f^\star} && \mathbb{N} \ar[dl]^{\id{N}} \\Y^\star \ar@{>->}[rr]^(.7){\lgh_{Y}} & & \mathbb{N}}
\end{minipage}

\noindent
\begin{minipage}[l]{.83\linewidth}
\epo*
\end{minipage}
\hfill 
\begin{minipage}[r]{.15\linewidth}
	\xymatrix@R=10pt{\mathcal{G}_1 \ar[r]_{h}\ar@{>->}[d]_{m}&\mathcal{G}_2\ar[d]^{n}\\\mathcal{G}_3\ar[r]^{z}&\mathcal{P}}
\end{minipage}
\begin{proof}\label{proof:epo}
	Let $\mathcal{P} $ be $(A, B, C, s, t, q)$ and let $(K_i, \pi_i^1, \pi_i^2)$ be the kernel pair of $q_i^\star \circ s_i$, for $i\in \{1, 2, 3\}$. Let also $(U, u_1, u_2)$ be the kernel pair of $q^\star \circ s$.

	\vspace{-.3cm}
	\noindent
	\begin{minipage}[l]{.7\linewidth}
			\setlength{\parindent}{1.5em}
		Consider now the cube on the right. By hypothesis and \Cref{rem:kleene} its left face is a pullback. Moreover its bottom face is not a pushout but it is still a pullback by the first point of \Cref{prop:regmono} and the third one of \Cref{prop:fact}.
	\end{minipage}\hfill 
	\begin{minipage}[r]{.4\linewidth}
	\xymatrix@C=10pt@R=2pt{&A_1\ar[dd]|\hole_(.65){q_1^\star \circ s_1}\ar[rr]^{h_E} \ar@{>->}[dl]_{m_E} && A_2 \ar[dd]^{q_2^\star \circ s_2} \ar@{>.>}[dl]_(.6){n_E} \\ A_3 \ar[dd]_{q_3^\star \circ s_3}\ar[rr]^(.7){z_E} & & A \ar[dd]_(.3){q^\star \circ s}\\&{Q_1^\star}\ar[rr]|\hole^(.65){h_Q^\star}\ar@{>->}[dl]_{m_Q^\star} && {Q_2^\star} \ar@{>->}[dl]^{n_Q^\star} \\{Q_3^\star} \ar[rr]^(.65){z_Q^\star} & & Q^\star}	\end{minipage}

	\noindent
	\begin{minipage}[l]{.2\linewidth}
		\xymatrix@R=15pt{K_1\ar[r]_{k_{h_E}}\ar@{>->}[d]_{k_{m_E}}&K_2\ar@{>->}[d]^{k_{n_E}}\\K_3\ar[r]^{k_{z_E}}&U}
	\end{minipage}\hfill
	\begin{minipage}[r]{.8\linewidth}
			\setlength{\parindent}{1.5em}
			By hypothesis and the first point of \Cref{prop:eqhyp_complete} the top face is a pushout. Thus we can apply \Cref{prop:kerset} to deduce that the square on the left is a pushout. 
			By computing we obtain
	\end{minipage}
\vspace{-.3cm}
	\begin{align*}
		q^\star \circ t \circ u_1 \circ f_n &= q^\star \circ t \circ n_E \circ \pi_2^1 = n_C^\star \circ q_2^\star \circ s_2 \circ \pi_2^1 = n_C^\star \circ q_2^\star \circ s_2 \circ \pi_2^2 =q^\star \circ t \circ u_2 \circ f_n \\
		q^\star \circ t \circ u_1 \circ f_k &= q^\star \circ t \circ k_E \circ \pi_3^1 = k_C^\star \circ q_3^\star \circ s_3 \circ \pi_3^1 = k_C^\star \circ q_3^\star \circ s_3 \circ \pi_3^2 =q^\star \circ t \circ u_2 \circ f_k
	\end{align*}
	We can therefore deduce that $q^\star \circ t \circ u_1 = q^\star \circ t \circ u_2$, and the claim follows.
\end{proof}
